Simulating High-Dimensional Multivariate Data

### 2021-02-24 index.Rmd
Revise index

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the bigsimr R Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when conducting Monte Carlo studies and evaluating quantitative methods. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions and a broad class of admissible Spearman correlation and Kendall $\tau$ matrices. We describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

@@

Massive high-dimensional data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools. This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.

@@

Another central issue lies in characterizing dependency between components in the high-dimensional random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).

<!-- 
Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-->

For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.

@@

Our study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. Next, we detail an illustrative low-dimensional example of basic and advanced use of the `bigsimr` R package. Then we proceed with Monte Carlo studies under various distributional assumptions to assess accuracy and computation times for increasingly large $d$. After the MC evaluations, we revisit our high-dimensional motivating example and employ our methods in commonplace statistical computing tasks. Finally, we discuss the method's utility, limitations, and future directions.

>>>>

--- 
title: "Simulating High-Dimensional Multivariate Data"
subtitle: "using the `Bigsimr` Package"
author:
  - Alfred G. Schissler
  - Edward J. Bedrick
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Tin Nguyen
  - Anna K. Panorska
  - Juli Petereit
  - Walter W. Piegorsch
  - Duc Tran
site: bookdown::bookdown_site
bibliography: ["bigsimr.bib", "packages.bib"]
abstract:  It is critical to realistically simulate data when employing Monte Carlo techinques and evaluating statistical methodology. Measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers resort to simulation designs that posit independence or perform arbitrary data transformations. This article introduces the `bigsimr` R interface to the Julia `Bigsimr` package. This high-package provides a flexible, scaleable procedure to simulate high-dimensional random vectors with arbitrary marginal distributions with specified Pearson, Kendall, or Spearman correlation matrices. It also contains additional high-performance features, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and compute the nearest correlation matrix. Finally, we demonstrate example applications enabled via `bigsimr` by applying these functions to a motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=878$ patients and dimension $d=1000$.
---

@@

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of analyzing such proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often with comparison to existing methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for high-dimensional data using existing algorithms and tools. This is particularly true when simulating massive *multivariate*, *non-normal* distributions, arising naturally in many fields of study.

@@

<!-- 
Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics.
-->
Another central issue lies in characterizing dependency between components in the high-dimensional random vector.
The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc).
For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix exactly is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependency, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependency measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology.
We do, however, provide Pearson matching with some caveats.

@@

Our study proceeds by providing background information, including a description of our motivating example application --- RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. Next, we detail an illustrative low-dimensional example of basic use of the `bigsimr` R package. Then we proceed with Monte Carlo studies under various bivariate distributional assumptions to assess accuracy. We conclude the Monte Carlo evaluations by summarizing the computation time for increasingly higher dimensional vectors. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely a MC joint probability calculation and HD-correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.
